The X-linked intellectual disability gene KDM5C is a sex-biased brake against germline programs in somatic lineages

### 2024-10-26 submission/manuscript.Rmd
add competing interests
add bioproject link
abstract update

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we examine how the X-linked intellectual disability gene lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), contributes to tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key human neurological phenotypes, aberrantly express many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes misexpressed in the _Kdm5c_-KO brain are unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

@@

## Data availability
### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject PRJNA1165148.

@@

# Author contributions
K.M.B. and S.I. conceived the study and designed the experiments. I.V. generated the ESC and exEpiLC WGBS data. K.M.B performed all data analysis and all other experiments. The manuscript was written by K.M.B and S.I. and edited by K.M.B, S.I., and I.V.

>>>>

A pivotal step in the evolution of multicellularity is the division labor among cellular lineages, including the the distinction of the germline from the soma. In the early embryo, genes that establish germline identity are repressed in somatic lineages through DNA and histone modifications. Failure to repress germline genes in somatic lineages is common signature of cancer and observed in select neurodevelopmental disorders, however it is currently unclear how factors like development and sex influence their repression and somatic misexpression. Here, we examine how cellular context influences the development of somatic tissue identity in mice with loss of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found KDM5C is a crucial regulator of tissue identity, as male _Kdm5c_ knockout (-KO) mice aberrantly express many liver, muscle, ovary, and testis genes within the brain. Late-stage spermatogenesis genes, but not somatic testicular genes, were highly enriched within the _Kdm5c_-KO brain, indicating an erosion of soma-germline boundary. Germline genes are typically decommissioned in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Characterizing germline gene misexpression in males and female mutants revealed germline gene repression is sexually dimorphic, with female EpiLCs requiring a higher dose of KDM5C to maintain germline gene suppression. Using a comprehensive list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation during ESC to EpiLC differentiation. However, late-stage spermatogenesis genes devoid of promoter CGIs can become expressed in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Together, these data demonstrate KDM5C's fundamental role in tissue identity and indicate that KDM5C acts as a brake against runaway activation of germline developmental programs in somatic lineages.

**notes**

- Distinguishing the germline from the soma is a key step in the evolution of multicellularity and sexual reproduction. 
- Germline gene repression is orchestrated by chromatin regulators and transcription factors. 
- Much of these discoveries have been made looking at the repression of key marker genes for germ cell specification in the early embryo,

- While the silencing mechanisms for genes that establish germline identity are well characterized, it is unclear if other types of germline genes, such as those involved in late oogenesis and spermatogenesis, employ the same silencing mechanisms and how silencing changes over the course of development. 
- In mammals, genes crucial for early germline specification gain repressive DNA and histone modifications in the early embryo. 
- Germline genes have not been comprehensively assessed as a whole

@@

## Data availability
### WGBS in wild-type and _Kdm5c_-KO ESCs and exEpiLCs
Raw fastq files are deposited in the Sequence Read Archive (SRA) https://www.ncbi.nlm.nih.gov/sra under the bioProject PRJNA1165148. https://www.ncbi.nlm.nih.gov/bioproject/PRJNA1165148

@@

# Author Contributions
K.M.B. and S.I. conceived the study and designed the experiments. I.V. generated the ESC and exEpiLC WGBS data. K.M.B performed all data analysis and all other experiments. The manuscript was written by K.M.B and S.I. and edited by K.M.B, S.I., and I.V.

# Declaration of Interest
S.I. is a member of the Scientific Advisory Board of KDM5C Advocacy, Research, Education & Support (KARES). Other authors declare no conflict of interest.
